AFQ-Insight

### 2019-11-15 reports/manuscript/sections/methods.tex
Add notebooks and requirements.txt
New figures (from OHBM poster). Lots of changes everywhere.

We tested AFQ-Insight in a classification setting using data from a previous study of the corticospinal tract profile in patients with amyotrophic lateral sclerosis (ALS)\cite{sarica2017corticospinal}. This study collected dMRI data from 24 ALS patients and 24 demographically matched healthy controls. In the regression setting, we used dMRI data from a previous study of properties of the white matter across the lifespan\cite{yeatman2014lifespan}, containing dMRI data from 76 subjects with ages 6-50. In both cases, the authors of these previous studies made their imaging data available through\ldots \adam{How? Through neurovault? Private correspondence?}.

The ``raw'' dMRI data from the ALS and lifespan maturation studies were then analyzed using the Automated Fiber-Tract Quantification tractometry pipeline\cite{yeatman2012tract} with the pyAFQ package\cite{pyAFQ} \adam{Or was it the MATLAB implementation?}. This creates tract profiles, in which diffusion measures are quantified and averaged along twenty major fiber tracts. These tract profiles, along with the phenotypical data we wish to explain or predict, form the input to AFQ-Insight. In a domain-agnostic machine learning context, the phenotypical data comprise the target variables while the tract profiles form the feature or predictor variables.

AFQ-Insight reads the target and feature data from comma separated value (CSV) files conforming to the AFQ-Browser data format\cite{yeatman2018browser} \adam{Should we cite the specific page in the online documentation} and represents them internally as \lstinline{DataFrame} objects from the pandas Python library\cite{mckinney2010data}. We leave imputation of the target data to the user and automatically impute missing data only in the feature matrix. Missing interior nodes are imputed using linear interpolation. For missing exteriod nodes, the user may choose between linear extrapolation and constant forward(back)-fill. The data imputation uses only values from adjacent nodes in the same white matter tract in the same subject so there is no danger of data leakage from other subjects.

Lastly, like most learning algorithms, AFQ-Insight benefits from standardization of the input dataset. Some diffusion metrics will have naturally larger variance than others and may therefore dominate the objective function and make the sparse group lasso estimator unable to learn from the lower variance metrics. For example, fractional anisotropy (FA) is bounded between zero and one and could be overwhelmed by an unscaled higher-variance metric like the mean diffusivity (MD). To prevent this, AFQ-Insight uses the \lstinline{StandardScaler} from scikit-learn\cite{scikit-learn}, to remove each feature's mean and scale it to unit variance. Scaling is performed separately within each cross-validation set's training or testing data to prevent leakage of information between the testing and training sets\cite{kaufman2012leakage}.

@@

@@

In addition to regularizing the coefficients in $\beta$, we also wish to capitalize upon our knowledge of the group structure of the tract profile features in $\mathbf{X}$. The tract-metric combinations form a natural grouping. For example, we expect that MD features within the left arcuate fasciculus will be correlated. Likewise for FA values within the right corticospinal tract (CST) and so on. This group structure is represented in Fig~\ref{fig:group-structure}, which depicts the linear model $y = \mathbf{X} \cdot \beta$. The target vector $y$ is sorted into patients (p) and controls (c), while the different colors encode the group structure of the feature matrix $\mathbf{X}$, with one group for each unique tract-metric combination.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.65\textwidth]{dMRI_group_structure.pdf}
    \caption{{\bf dMRI group structure.}
        The phenotypical target data and tractometric features can be organized into a linear model, $y = \mathbf{X} \cdot \beta$. Here, the target array $y$ is suggestively sorted into patients (p) and controls (c), while the feature matrix $\mathbf{X}$ is color-coded to reveal a natural group structure. In this example, the left (orange) group contains $k$ features from the inferior fronto-occipital fascicle (IFOF), the middle (green) group contains $k$ features from the corpus callosum, and the right (blue) group contains $k$ features from the uncinate. The coefficients in $\beta$ follow the same natural grouping. Fascicle images taken with permission from Ref~\cite{yeatman2012tract}.
    }
    \label{fig:group-structure}
\end{figure}

Thus, we seek a regularization approach that will fit a linear model with anatomically-grouped covariates, where we expect to observe sparse coefficients at both the group level and the within-group level. The sparse group lasso (SGL) is a penalized regression technique that satisfies exactly these criteria\cite{simon2013sparse}. It solves for a coefficient vector $\widehat{\beta}$ that satisfies
\begin{equation}
    \widehat{\beta} = \min_\beta \frac{1}{2}
    \norm*{y - \displaystyle \sum_{\ell = 1}^{m} \mathbf{X}^{(\ell)} \cdot \beta^{(\ell)}}_2^2
    + \lambda_1 \displaystyle \sum_{\ell = 1}^{m} \sqrt{p_\ell} \norm*{\beta^{(\ell)}}_2
    + \lambda_2 \norm*{\beta}_1,
    \label{eq:sgl}
\end{equation}
where $m$ is the number of anatomical groups, $\mathbf{X}^{(\ell)}$ is the submatrix of $\mathbf{X}$ corresponding to group $\ell$, $\beta^{(\ell)}$ is the coefficient vector for group $\ell$ and $p_\ell$ is the length of $\beta^{(\ell)}$. The first term is the mean square error loss, $L_{mse}$, as in the standard linear regression framework. The second and third terms encourage groupwise sparsity and overall sparsity, respectively. If $\lambda_1 = 0$ and $\lambda_2 = 1$, the SGL reduces to the traditional lasso\cite{tibshirani1996regression}. Conversely, if $\lambda_1 = 1$ and $\lambda_2 = 0$, the SGL reduces to the group lasso\cite{yuan2006model}.

When the phenotypical target variable are categorical, as in the case of predicting a clinical diagnosis, the SGL is readily adapted to logistic regression, where the probability of a target variable belonging to an arbitrary defined ``true'' class is the logistic function of the result of the linear model,
\begin{equation}
    p(y = 1) = \frac{1}{1 + \exp(\mathbf{X}\cdot \beta)},
    \label{eq:logit}
\end{equation}
or equivalently, the mean squared error loss function in Eq~\eqref{eq:sgl} is replaced with the cross-entropy loss, which for binary classification is the negative log likelihood of the SGL classifier giving the ``true'' label:
\begin{equation}
    L_{mse} \rightarrow L_{\log} = -\left(y \log(p) + (1 - y) \log(1 - p)\right).
    \label{eq:logloss}
\end{equation}

@@

>>>>

Two different previously-published datasets were used here:

\begin{enumerate}

\item Diffusion MRI data from a previous study of properties of the white matter
across the lifespan\cite{yeatman2014lifespan}, containing dMRI data from 76
subjects with ages 6-50. These data were measured in a GE Discovery 750 3T MRI
scanner at the Stanford Center for Cognitive and Neurobiological Imaging. The
Stanford University IRB approved the procedures of this study. Informed consent
was obtained from each adult participant, and assent for participation was
provided by by parents/guardians for children. Voxel resolution was 2x2x2 $mm^3$
96 non-colinear directions of measurement were measured with a b=2,000 $s/mm^2$
and 30 non-colinear directions of measurement were measured with a b=1,000
$s/mm^2$. Data was preprocessed to correct for subject motion and the diffusion
tensor model \cite{basser1994mr} was fit in every voxel, using a robust fit
\cite{chang2005restore}. These data were acquired using a dual spin echo
sequence, in which there is sufficient time for eddy currents to subside between
the application of the gradients and the image acquisition, so no eddy current
correction was appliced. We will refer to this dataset as WH.

\item Diffusion MRI from from a previous study of the corticospinal tract
profile in patients with amyotrophic lateral sclerosis
(ALS)\cite{sarica2017corticospinal}, containing data from 24 ALS patients and 24
demographically matched healthy controls. These data were measured in a GE
Discovery 750 3T MRI scanner at the Institute of Bioimaging and Molecular
Physiology in Catanzaro. Informed consent was provided as approved by the
Ethical Committee of the University ``Magna Graecia'' of Catanzaro. Voxel
resolution was 2x2x2 $mm^3$ and 27 non-colinear directions were measured with a
b=1,000 $s/mm^2$. Here, both motion correction and eddy current correction were
applied before modeling the diffusioon tensor in each voxel. We will refer to
this dataset as ALS.

\end{enumerate}

Data from both of these studies was processed in a similar manner, using the
Matlab Automated Fiber Quantification toolbox (AFQ) \cite{yeatman2012tract}:
streamlines representing fascicles of white matter nerve fibers were generated
using a determinstic tractography algorithm that follows the prinicpal diffusion
direction of the diffusion tensor in each voxel (STT \cite{basser2000vivo}).
Major tracts were identified using multiple criteria: streamlines are selected
as candidates for inclusion in a bundle of streamlines that represents a tract
if they pass through known inclusion ROIs and do not pass through exclusion ROIs
\cite{wakana2007reproducibility}. In addition, a probabilistic atlas is used to
exclude streamlines which are unlikely to be part of a tract \cite{Hua2008-sh}.
Each streamline is resampled to 100 nodes and the robust mean at each location
is calculated by estimating the 3D covariance of the location of each node and
excluding streamlines that are more than 5 standard deviations from the mean
location in any node. Finally, a tract profile of tissue properties in each
tract was created by interpolating the value of MRI maps of these tissue
properties to the location of the nodes of the resampled streamlines designated
to each tract. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the mahalnobis
distance of the node from the average  of that node across streamlines. This
means that streamlines that are more representative of the tract contribute more
to the tract profile, relative to streamlines that are on the edge of the tract.

This process creates tract profiles, in which diffusion measures are quantified
and averaged along twenty major fiber tracts. Here, we use only the mean
diffusivity (MD) and the fractional anisotropy (FA) of the diffusion tensor, but
additional dMRI-based maps or maps based on other quantitative MRI measurements
can also be used. These tract profiles, along with the phenotypical data we wish
to explain or predict, form the input to the SGL algorithm. In a domain-agnostic
machine learning context, the phenotypical data comprise the target variables
while the tract profiles form the feature or predictor variables (See
Fig~\ref{fig:group-structure}).

@@

Before fitting a model to the data, imputation and standardization are
performed. Missing node values (e.g., in cases where AFQ designates a node as
not-a-number) are imputed via linear interpolation. Care is taken not to
interpolate across the boundaries between different tracts. Some diffusion
metrics will have naturally larger variance than others and may therefore
dominate the objective function and make the SGL estimator unable to learn from
the lower variance metrics. For example, fractional anisotropy (FA) is bounded
between zero and one and could be overwhelmed by an unscaled higher-variance
metric like the mean diffusivity (MD). To prevent this, we use the
\lstinline{StandardScaler} from scikit-learn\cite{scikit-learn}, to remove each
feature's mean and scale it to unit variance. Scaling is performed separately
within each cross-validation set's training or testing data to prevent leakage
of information between the testing and training sets\cite{kaufman2012leakage}.

@@

In addition to regularizing the coefficients in $\beta$, we also wish to
capitalize on our knowledge of the group structure of the tract profile features
in $\mathbf{X}$. The tract-metric combinations form a natural grouping. For
example, we expect that MD features within the left arcuate fasciculus will be
correlated. Likewise for FA values within the right corticospinal tract (CST)
and so on. This group structure is represented in Fig~\ref{fig:group-structure},
which depicts the linear model $y = \mathbf{X} \cdot \beta$. Thus, we seek a
regularization approach that will fit a linear model with anatomically-grouped
covariates, where we expect to observe sparse coefficients at both the group
level and the within-group level. The sparse group lasso (SGL) is a penalized
regression technique that satisfies exactly these
criteria\cite{simon2013sparse}. It solves for a coefficient vector
$\widehat{\beta}$ that satisfies
\begin{equation}
    \widehat{\beta} = \min_\beta \frac{1}{2}
    \norm*{y - \displaystyle \sum_{\ell = 1}^{m} \mathbf{X}^{(\ell)} \cdot \beta^{(\ell)}}_2^2
    + \lambda_1 \displaystyle \sum_{\ell = 1}^{m} \sqrt{p_\ell} \norm*{\beta^{(\ell)}}_2
    + \lambda_2 \norm*{\beta}_1,
    \label{eq:sgl}
\end{equation}

where $m$ is the number of groups, $\mathbf{X}^{(\ell)}$ is the submatrix of
$\mathbf{X}$ corresponding to group $\ell$, $\beta^{(\ell)}$ is the coefficient
vector for group $\ell$ and $p_\ell$ is the length of $\beta^{(\ell)}$. The
first term is the mean square error loss, $L_{mse}$, as in the standard linear
regression framework. The second and third terms encourage groupwise sparsity
and overall sparsity, respectively. If $\lambda_1 = 0$ and $\lambda_2 = 1$, the
SGL reduces to the traditional lasso\cite{tibshirani1996regression}. Conversely,
if $\lambda_1 = 1$ and $\lambda_2 = 0$, the SGL reduces to the group
lasso\cite{yuan2006model}.

When the phenotypical target variable are categorical, as in the case of
predicting a clinical diagnosis, the SGL is readily adapted to logistic
regression, where the probability of a target variable belonging to an arbitrary
defined ``true'' class is the logistic function of the result of the linear
model,

\begin{equation}
    p(y = 1) = \frac{1}{1 + \exp(\mathbf{X}\cdot \beta)},
    \label{eq:logit}
\end{equation}

or equivalently, the mean squared error loss function in Eq~\eqref{eq:sgl} is
replaced with the cross-entropy loss, which for binary classification is the
negative log likelihood of the SGL classifier giving the ``true'' label:

\begin{equation}
    L_{mse} \rightarrow L_{\log} = -\left(y \log(p) + (1 - y) \log(1 - p)\right).
    \label{eq:logloss}
\end{equation}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.65\textwidth]{dMRI_group_structure.png}
    \caption{{\bf dMRI group structure.}
        The phenotypical target data and tractometric features can be organized
    into a linear model, $y = \mathbf{X} \cdot \beta$. The feature matrix
    $\mathbf{X}$ is color-coded to reveal a natural group structure: the left
    (orange) group contains $k$ features from the inferior fronto-occipital
    fascicle (IFOF), the middle (green) group contains $k$ features from the
    corpus callosum, and the right (blue) group contains $k$ features from the
    uncinate. The coefficients in $\beta$ follow the same natural grouping.
    Fascicle images taken from Ref~\cite{yeatman2012tract}.}
    \label{fig:group-structure}
\end{figure}

\subsection*{Metaparameter optimization}

To select the values of the metaparameters $\lambda$

@@

\subsection*{Software implementation}

The full software implementation of the SGL approach presented here is available
in a Python software package called AFQ-Insight, which is developed publicly in
\url{https://github.com/richford/afq-insight}. The version of the code used to
produce the results herein is also available in XXX Zenodo DOI. AFQ-Insight
reads the target and feature data that has been processed by AFQ from comma
separated value (CSV) files conforming to the AFQ-Browser data
format\cite{yeatman2018browser} and represents them internally as
\lstinline{DataFrame} objects from the pandas Python
library\cite{mckinney2010data}. The software provides different options for
imputing missing data in the feature matrix. Missing interior nodes are imputed
using linear interpolation. For missing exterior nodes, the user may choose
between linear extrapolation and constant forward(back)-fill. Imputation uses
only values from adjacent nodes in the same white matter tract in the same
subject so there is no danger of data leakage from other subjects.

### 2019-11-15 reports/manuscript/sections/results.tex
Add style file
New figures (from OHBM poster). Lots of changes everywhere.

Using data from a previous study of the corticospinal tract profile and ALS\cite{sarica2017corticospinal}, we tested the performance of AFQ-Insight in a classification setting. The previous study predicted ALS status with a mean accuracy of 80\% using a random forest algorithm based on a priori selection of features within the corticospinal tract. AFQ-Insight delivers competitive predictive performance (mean 84\% accuracy, 0.93 ROC AUC) without the need for a priori feature engineering. The results of the classification prediction are shown in Fig~\ref{fig:class-results}, with ``ground-truth'' ALS status separated into columns, and predicted ALS status encoded by color. In addition to this classification performance, AFQ Insight also identifies the white matter tracts most important for the ALS classification. The relative importance of white matter features is captured in the $\beta$ coefficients from Eq~\eqref{eq:sgl}. Fig~\ref{fig:class-beta} depicts these coefficients ``unfolded'' across the brain. The $x$-axis organizes white matter tracts starting with left distal tracts, progressing toward the center of the brain and then outward toward the right distal tracts. Fig~\ref{fig:class-beta} shows that AFQ-Insight selects FA metrics in the right corticospinal tract as most important to ALS classification, confirming previous findings\cite{van2011upper, toosy2003diffusion, sarica2014tractography, sage2007quantitative, sage2009quantitative, karlsborg2004corticospinal, ellis1999diffusion, cosottini2005diffusion, ciccarelli2009investigation, abe2010voxel} and identifying exactly the portions of the brain that were selected \emph{a priori} in the previous study from which we collected our data\cite{sarica2017corticospinal}.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.65\textwidth]{classification_results.png}
    \caption{{\bf Prediction of ALS status.}
        Classification probabilities for each subject's ALS diagnosis. Controls are on the left while patients are on the right. Predicted controls are in blue and predicted patients are in red. Thus, false positive are represented as red dots on the left, while false negatives are represented as blue dots on the right. AFQ-Insight achieves 84\% accuracy with an ROC area under the curve of 0.93.
    }
    \label{fig:class-results}
\end{figure}

@@

In a regression setting, we attempted to accurately predict ``brain age''.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.65\textwidth]{regression_results.png}
    \caption{{\bf Prediction of brain age.}
    }
    \label{fig:class-results}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.95\textwidth]{regression_unfolded_beta.png}
    \caption{{\bf Feature importance for brain age.}
    }
    \label{fig:class-beta}
\end{figure}

>>>>

The WH dataset used for continuous regression contains data from 76 healthy
subjects, ranging between 6 years and 50 years of age
\cite{yeatman2014lifespan}. SGL fit to tractometry-extracted features in these
subjects accurately predicts the age of the subjects. The model reaches a mean
absolute error of 3.6 years (Figure \ref{fig:regress-results}), which is lower
than the results of a recent study that predicted age in a large sample, based
on TBSS-extracted diffusion features \cite{Richard2018-ux}. The model weights
in this case are distributed over many different tracts and dMRI tissue
properties (Figure \ref{}) This demonstrates that SGL is not coerced to produce overly sparse
results when a more accurate model requires a dense selection of features.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.65\textwidth]{regression_residuals.png}
    \caption{{\bf Prediction of brain age.}
    }
    \label{fig:regress-results}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.95\textwidth]{regression_beta_annotated.pdf}
    \caption{{\bf Feature importance for brain age.}
    }
    \label{fig:regress-beta}
\end{figure}

Using data from a previous study of the corticospinal tract profile and
ALS\cite{sarica2017corticospinal}, we tested the performance of SGL in a
classification setting. The previous study predicted ALS status with a mean
accuracy of 80\% using a random forest algorithm based on a priori selection of
features within the corticospinal tract. AFQ-Insight delivers competitive
predictive performance (mean 93\% $\pm$ 2\% accuracy, 0.978 $\pm$ 0.006 ROC
AUC) without the need for a priori feature engineering. The results of the
classification prediction are shown in Fig~\ref{fig:class-results}, with
``ground-truth'' ALS status separated into columns, and predicted ALS status
encoded by color. In addition to this classification performance, AFQ Insight
also identifies the white matter tracts most important for the ALS
classification. The relative importance of white matter features is captured in
the $\beta$ coefficients from Eq~\eqref{eq:sgl}. Fig~\ref{fig:class-beta}
depicts these coefficients ``unfolded'' across the brain. The $x$-axis organizes
white matter tracts starting with left distal tracts, progressing toward the
center of the brain and then outward toward the right distal tracts.
Fig~\ref{fig:class-beta} shows that AFQ-Insight selects FA metrics in the right
corticospinal tract as most important to ALS classification, confirming previous
findings\cite{van2011upper, toosy2003diffusion, sarica2014tractography,
sage2007quantitative, sage2009quantitative, karlsborg2004corticospinal,
ellis1999diffusion, cosottini2005diffusion, ciccarelli2009investigation,
abe2010voxel} and identifying exactly the portions of the brain that were
selected \emph{a priori} in the previous study from which we collected our
data\cite{sarica2017corticospinal}.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.65\textwidth]{classification_probs.png}
    \caption{{\bf Prediction of ALS status.}
        Classification probabilities for each subject's ALS diagnosis. Controls are on the left while patients are on the right. Predicted controls are in blue and predicted patients are in red. Thus, false positive are represented as red dots on the left, while false negatives are represented as blue dots on the right. AFQ-Insight achieves 84\% accuracy with an ROC area under the curve of 0.93.
    }
    \label{fig:class-results}
\end{figure}

@@
